---
title: "W3D2: Mutating and selecting data"
format: typst
---

# Goals for today

- Learn how to use the `mutate()` function in `dplyr` to calculate new variables for our data that we can use for ranking, filtering, and other operations.

::: callout-note
This activity focuses on data wrangling rather than visualization, so we will ignore data labels this time.
:::

------------------------------------------------------------------------

# Setup

You know the drill: load the packages and data!

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(patchwork)

chetty <- read_excel("./data/Lab1_Chetty_2014.xlsx", skip = 1)
```

Then create a subset of `chetty` by creating a new data frame consisting of only the variables we need for our analysis:

```{r}
chetty <- chetty |>
  select(cz_name, 
         state, 
         pop_2000, 
         abs_mobility, 
         prob_q1q5,
         starts_with("frac"),
         hhi_percap,
         urban, 
         region)
```

## Mutating data

The `mutate()` function lets us calculate new variables from existing ones.

We start by naming the variable we want to create, then set it equal (=) to the output of some operation, like arithmetic or an R function:

```{r}

```

## Usefulness of mutate

`mutate()` is fantastic for supplementing our data with new calculations. In this course, we will use it often to calculate summary statistics like ranks, means, and percentiles that help us understand our data better.

We can also use `mutate()` to clean our data, concatenate variables together, split them apart, or perform various other transformations that are critical to data analysis.

## New functions today

|                  |                                                    |
|------------------|----------------------------------------------------|
| `log2()`         | Computes the logarithm of a given column in base 2 |
| `percent_rank()` | Computes a percentile                              |
| `ntile()`        | Groups data by n buckets of equal size             |
| `min_rank()`     | Ranks values, and skips when there is a tie        |
| `cumsum()`       | “Cumulative sum” - computes a running total        |
| `relocate`       | Move specified columns to the front                |

------------------------------------------------------------------------

# Exercise 1

## Step 1

Make new variables for each of these quantities:

- `rel_pop` = the number of people in each CZ who consider themselves to be religious
- `log2income` = the log base 2 of the per capita household income (you can use the built-in `log2()` function for this)
- `not_black` = the proportion of people who are **not** Black/African American

Then use `select()` to only include the CZ name, state, and the three new variables you just created.

```{r}

```

## Step 2

It's not always necessary to save transformed data as a new data frame, especially when you're still experimenting and you're not sure you have the output you need. However, once you have confirmed your code works and gives you the right output, you can overwrite the original subset data frame using the assignment operator, like so: `df <- df |> blah blah blah...`

Remember: to see the results, you can wrap the whole statement inside parentheses.

```{r}

```

::: callout-important
Note how we overwrote our data frame, but `mutate()` allowed us to simply tack on the new variables
:::

## Detour

**Why are we calculating the log base 2 of per capita household income??**

Income data from geographic areas can often be right-skewed, meaning that a few high-income households can have an outsize influence on summary statistics like averages. Those outliers can distort our analysis, so a logarithmic transformation like this helps to mitigate the presence of the outliers, making our data less vulnerable to extreme values and leading to more robust statistical inferences.

Specifically, taking the log base 2 of income emphasizes **relative changes** in our income variable, rather than absolute differences.

For example, the absolute difference between \$1,000 and \$2,000 is \$1,000, correct? And the absolute difference between \$10,000 and \$20,000 is \$10,000, correct? In both situations, the income has doubled, but the absolute change is much different.

With that in mind, let's see what happens when we take the log base 2 in both situations:

$$\log_2(2000) - log_2(1000) = log_2(\frac{2000}{1000}) = log_2(2) = 1$$

$$\log_2(20000) - log_2(10000) = log_2(\frac{20000}{10000}) = log_2(2) = 1$$

The outputs are exactly the same! This means doubling income is represented by a constant increase in the $log_2$ value, regardless of the initial income value. In short, that $log_2$ value helps us quantify the relative change in incomes.

This process of adjusting or re-scaling values to minimize the effect of the outliers is referred to as **normalization**. We have "normalized" our income variable to emphasize relative change over absolute change, and produce a distribution with less skewness.

To visualize the difference between our original per capita income variable and our new normalized variable, let's generate a boxplot for each. We'll use the excellent `patchwork` package to stitch both plots together into a single output:

```{r}

```

This also helps us visually see the results of the log base 2 transformation: it compresses the higher values and expands the lower values, creating a distribution that is more symmetrical around the mean. This is closer to a **normal distribution**, one of the biggest concepts in probability. Many of the most important and common statistical methods for analyzing numerical data (like income) assume the data are normally distributed, so doing this kind of transformation can improve the accuracy and reliability of our analysis.

::: callout-note
**Did you know?**: Normal distributions are naturally-occurring. They can be used to describe many biological random variables, including people's height, weight, blood pressure, and IQ scores.
:::

------------------------------------------------------------------------

# Exercise 2

## Step 3

Make new variables that indicate which percentile and which quantile the per capita household income is in among all CZs:

- `ptile_hhipercap`: use the `percent_rank()` function here.
- `qtile_hhipercap`: use the `ntile()` function here, with the second argument being used to specify that we are looking for quantiles (4).

Note: we can use the `everything()` function to make sure we include all of our original variables in our output.

```{r}
  
```

Calculating these percentiles and quantiles helps us locate specific CZ's along the overall distribution.

## Step 4

Create a new variable that returns TRUE whenever a CZ is ranked in the top 50 commuting zones by population. You can use the `min_rank()` function here with the `pop_2000` variable to accomplish this.

```{r}

```

## Step 5

OK, let's try something more complicated that demonstrates how we can chain these transformations together. Make a data frame named `worst_mobility` containing:

- The 41st to 50th worst absolute mobility scores;
- among the 50 biggest CZs (measured by population);
- sorted in that order; and
- displaying only the rank, CZ name, absolute mobility, and population

```{r}

```

There is a second way of doing this:

```{r}

```

## Step 6

Lastly, add a variable to the data frame you made in Step 5 that tracks the **running total** of people in these bottom 10 of the 50 largest CZs.

::: callout-hint
Running totals can be calculated using the `cumsum()` function.
:::

```{r}

```

------------------------------------------------------------------------

::: callout-important
You've reached the end of the exercise!
:::
