---
title: "W7D1: Mutating joins"
format: typst
---

# Goals for this class

- Learn how to use mutating joins like `left_join()` and `inner_join()` to merge two tables or data frames together based on attributes they have in common.
- Explore data about worldwide gender income disparities.

# Joins!

It is common in data science & data engineering that we need to solve a problem or answer a question using data that comes from **multiple** data sources.

For example, we can combine economic data about countries from the [World Bank](https://databank.worldbank.org/){target="_blank"} with inequality data from the [World Inequality Database](https://wid.world/data/){target="_blank"}, to try to answer questions about how economic growth in certain countries may relate to advancements in gender equality.

## Two main types of joins

**Mutating joins** (today)

Columns from Table 2 are added to columns from Table 1, matching observations between the two data frames using a shared key column.

**Filtering joins** (next class)

Columns from Table 2 are used to filter Table 1 based on matching observations.

## Key columns 🔑

[Per R4DS](https://r4ds.hadley.nz/joins.html#keys){target="_blank"}: "Every join involves a pair of keys..."

A **Primary Key (PK)** is used to uniquely identify each observation in a data frame.

A **Foreign Key (FK)** is a key column on one table that corresponds to the PK in another table, creating an implicit *relationship* between the two tables.

# Four Types of Joins

## Left join

Get all rows from `table1` plus any matching rows from `table2`

## Right join

Get all rows from `table2` plus any matching rows from `table1`

## Full outer join

Get all rows from both `table1` and `table2`, even if they don't have matches.

## Inner join

Get only matching rows from **both** tables.

::: callout-warning
Confusion has joined the chat!
:::

# In-Class Exercise

## Setup

You know the drill: load your packages!

```{r setup, message = F, warning=F}
library(tidyverse)
library(readxl)
library(ggrepel)
```

And now let's load our data.

The main dataset for today's lesson comes to us from the [World Inequality Database (WID)](https://wid.world/data/){target="_blank"}, a product of the [World Inequality Lab](https://inequalitylab.world/en/){target="_blank"}, which is itself a research center based out of the Paris School of Economics and UC Berkeley. The WIL's mission is to study global inequality and promote evidence-based public policies centered on issues of social, economic and environmental justice.

We are importing two data frames into RStudio. The first one, `countries`, is a simple list of world countries and their regions and subregions. The second one, `labor`, contains yearly data about female labor income share (see below).

```{r, message=FALSE}

countries <- read_delim("./data/WID_countries.csv", delim=";") |>
  filter(!is.na(region)) |>
  select(
    alpha2,
    country = shortname,
    region,
    subregion = region2,
  )

labor <- read_excel(
  path = "./data/WID_female_inc_share.xlsx",
  col_names = c("country","indicator","indicator_code","year","value")
  ) |>
  select(country, year, value)
```

## FLIS

"Female labor income share" (FLIS) is defined as the total labor income received by women **divided by** the total labor income of the population. In other words, it's a percentage representing the share of global labor income (including wages and salaries) earned by women.

$$
FLIS = \frac{\text{Labor income received by women}}{\text{Total labor income}}
$$

For more detail on the WID's methodology behind this metric, check out their [technical notes](https://wid.world/www-site/uploads/2024/12/WorldInequalityLab_TechnicalNote_2024_13_Female-Labor-Income-Share_Final.pdf){target="_blank"}.

# Exercises

## Part 1

### Left join

Remember what [§ 19.3.2 of R4DS](https://r4ds.hadley.nz/joins.html#specifying-join-keys){target="_blank"} tells us:

> By default, `left_join()` will use all variables that appear in both data frames as the join key, the so called **natural join**.

We can see a natural join in action when we attempt our first left join by **not** explicitly specifying a join key, like so. The output of the `left_join()` function will include a quick hint telling us what column(s) it chose to join our two tables on:

```{r}

```

::: callout-tip
Implicit (natural) joins can be OK, but be careful with them. The R4DS excerpt linked above includes a very good example of how one of these implicit joins can get you into trouble. **In general, I highly recommend you use explicit join syntax in your code, even when it seems unnecessary**.
:::

To do an explicit join, simply specify the `by =` argument inside the `left_join()` function:

```{r}

```

Note how this produces the exact same output as the previous example, but anybody who comes after us to try to reproduce our code has a clearer idea of **how** we joined these two tables because we explicitly specified the relationship.

### Inner join

Let's also run through a quick example of an **inner join** to see how we can find all rows that match in both tables:

```{r}

```

If we want to locate the difference of 31 records between our `left_join()` and `inner_join()` examples, we can simply take our left-joined data frame and filter to locate all the NA values:

```{r}

```

## Part 2

Now that we've explored the difference between `left_join()` and `inner_join()`, let's take a crack at visualizing some yearly time series data that illuminates gender income disparities between different regions of the world.

### Grouping by region

First, we'll begin by grouping our data by region **and year**. Let's create a new data frame named `labor_by_region` which aggregates all this for us, and let's calculate the mean FLIS:

```{r, message=F}

```

### Line chart 1 (by region)

Now let's take our grouped data frame and load it into a ggplot visual. In particular, let's use `geom_line()` to create a line chart that shows us how the FLIS measure has changed on a yearly basis. To do this, we can put our year variable on the X axis.

To streamline this lesson a bit, I've staged a ggplot2 template function for you which accepts three arguments:

1.  `data` = a data frame
2.  `groupvar` = the name of whatever categorical variable we want to group by in our plot
3.  `value` = the name of whatever numerical variable we want to visualize

Let's begin by running this code to add the template function to our workspace:

```{r}
FLIS_chart <- function(data, groupvar, value) {
  warning("This is not an error. This is a joke.")
  data |> 
  ggplot(aes(x = year, y = {{ value }},color = {{ groupvar }})) + 
    geom_line(linewidth = 1) + 
    geom_hline(yintercept = 0.5,
               color = "blue",
               linewidth  = 2,
               linetype   = "dotted") + 
    labs(title = "Female Labor Income Share",
         caption = "Source: World Inequality Database",
         x = "Year",
         y = "Average FLIS"
    ) + 
    geom_text_repel(aes(label = ifelse(year == max(year), {{ groupvar }}, NA)), 
      nudge_x       = 1,    # Push labels slightly right
      direction     = "y",  # Allow movement only in the y-direction
      hjust         = 0, 
      box.padding   = 0.3,  # Adjust padding to avoid overlaps
      segment.color = NA    # Hide repelling lines
    ) +
    theme_minimal() +
    scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
    scale_y_continuous(labels = scales::percent) + 
    theme(legend.position = "none")
}
```

With that done, let's input our `labor_by_region` data frame to create a line chart with our `FLIS_chart()` function, putting the `mean_FLIS` variable on the Y axis, and grouping by **region**:

::: callout-note
By tacking a new `labs()` function onto the end of this, we can also add new labels to our ggplot2 object, or we can override the labels that were defined in our function!
:::

```{r}

```

### Grouping by subregion

**Which sub-regions in Asia have the highest gender parity?** If we want to answer this question, we need to aggregate our FLIS data at a more precise or granular level. Specifically, we need to determine the mean FLIS value for each subregion. So let's repeat the same process we did earlier, but with a slight variation in our code, to produce a new data frame named `labor_by_subregion`.

```{r, warning = F, message = F}

```

### Grouping by country

**Which countries in East Asia have the highest gender parity?** If we want to answer this question, we need to aggregate our FLIS data at a more precise or granular level. Specifically, we need to determine the mean FLIS value for each subregion. So let's repeat the same process we did earlier, but with a slight variation in our code, to produce a new data frame named `labor_by_subregion`.

```{r, warning = F, message = F}


```

## Part 3

Lastly, let's join even more data by bringing in the World Bank data that we used in one of our earliest exercises, creating a new data frame named `worldbank`:

```{r, message = FALSE}

```

And now let's join this new data frame to one of our earlier joined data frames by joining on some column(s) they share in common.

First, let's try it with a natural (implicit) join to see what columns R thinks we should join on:

```{r}

```

::: callout-tip
Note how R replies back with this message when confronted with the natural key join:

```         
Joining with `by = join_by(country, year, region)`
```

It looked at both tables and found three columns with shared names, but this poses multiple problems for us:

- Country names and region names can be different depending on the source of the data.
- Our World Bank data frame has data for the year 2022, but our WID data has data from 1990 to 2024.

This is all evident in the fact that our code above does not return any values for the columns `alpha2`, `subregion`, or `value`, all of which come from our World Inequality data. This tells us that R was not able to find any matches based on these criteria.
:::

With these differences in mind, it's probably a good idea for us to be more explicit in our join syntax and use a column that we can probably tell is **standardized** between our World Bank data and our WID data:

```{r}

```

------------------------------------------------------------------------

::: callout-important
You've reached the end of the class!
:::
