---
title: "Homework 07"
format: typst
---

# Intro

This assignment focuses on **relational data** and **joins** using the `dplyr` package.

In contrast with most of the datasets in this class, which focus on weighty topics of social and economic justice, this assignment will divert us to a lighter and more fun subject: sports analytics! You’ll work with datasets from the `Lahman` package, which contains historical baseball data, including information about players, teams, batting, fielding, pitching, and awards. Your task is to use `dplyr` joins to combine and analyze these datasets.

As you work through this homework assignment, you’ll practice using **mutating joins** (e.g., `left_join()`) and **filtering joins** (e.g., `semi_join()` and `anti_join()`). These skills are essential for working with real-world datasets that are often spread across multiple tables.

**Note:** For this assignment, make sure to add appropriate labels to each visualization.

Happy coding!

# Setup

Because we're dealing with a new package that does not come pre-installed in JupyterHub, you will need to run the following command in your RStudio console:

::: callout-note
You can attempt to run this in a Quarto code block, but I don't recommend it. Use the Console instead for the best results!
:::

```{r, eval=FALSE}
install.packages("Lahman")
```

Next, run the setup block below to load the packages you need, including our new Lahman package.

```{r setup, warning=F, message=F}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Lahman)
```

# Data preview

Run the code block below to preview the datasets from the Lahman package. These datasets include `Batting`, `People`, `Teams`, `Salaries`, `Pitching`, `Fielding` and `AwardsPlayers`. Use `head()` or `View()` to explore the data.

```{r}
# Preview datasets
head(Batting)
head(People)
head(Teams)
head(Salaries)
head(Pitching)
head(Fielding)
head(AwardsPlayers)
```

# Exercises

## Question 1

Use a **mutating join** to add player information from the `People` dataset to the `Batting` dataset. Save the result in a new dataset called `batting_with_names`. Ensure the new dataset retains all rows from `Batting`, even if there is no matching player information.

::: callout-tip
Use `left_join()` to combine Batting and People on the `playerID` column.
:::

```{r}

```

## Question 2

Use a **filtering join** to find all unique players in the `batting_with_names` dataset who have won an award (as recorded in the `AwardsPlayers` dataset). Save the result in a new dataset called `batting_award_winners`. Include only the columns `playerID`, `teamID`, `nameFirst`, and `nameLast`.

::: callout-tip
Use `semi_join()` to filter `batting_with_names` based on matching rows in `AwardsPlayers` using the `playerID` column.
:::

```{r}

```

### Question 3

Use a **filtering join** to find all unique players players in the `batting_with_names` dataset who have not won an award. Save the result in a new dataset called `batting_no_awards`. Include only the columns `playerID`, `teamID`, `nameFirst`, and `nameLast`.

::: callout-tip
Use `anti_join()` to filter `batting_with_names` based on non-matching rows in `AwardsPlayers` using the `playerID` column.
:::

```{r}

```

### Question 4

Use a **mutating join** to add team information from the `Teams` dataset to the `batting_with_names` dataset. Filter the dataset to include only players who have had a World Series win (indicated by "Y" in the WSWin column). Save the result in a new dataset called `players_ws_wins` Include the columns `playerID`, `nameFirst`, `nameLast`, `yearID`, `teamID`, `W`, `L`, and `WSWin`. Then, create another dataframe called `top_12_ws_wins` using the `players_ws_wins` dataset, which contains the top 12 MLB players with the most World Series wins. Include the columns `playerID`, `nameFirst`, `nameLast`, and `ws_wins` (the count of World Series wins).

::: callout-tip
Use `left_join()` to combine `batting_with_names` and `Teams` on the `teamID` and `yearID` columns. Use `group_by()` and `summarize()` to count World Series wins for each player. Use `arrange()` and `slice()` to find the top 12 players with the most World Series wins.
:::

```{r}

```

### Question 5

Calculate the total number of career home runs (HR) for each player in the `Batting` dataset. Use a **mutating join** to add player names from the `People` dataset to the result. Save the final dataset as `players_career_hr`. Include the columns `playerID`, `nameFirst`, `nameLast` and sort the players by total number of career HRs in descending order.

```{r}

```

### Question 6

Use **multiple joins** to create a dataset that combines information from the `Batting`, `Pitching`, and `People` datasets. Save the result in a new dataset called `players_batting_pitching`. Include the columns `playerID`, `nameFirst`, `nameLast`, `yearID`, `teamID`, `HR` (home runs from Batting), and `SO` (strikeouts from Pitching). Make sure the dataset only includes players who have both batting and pitching statistics in the same year. Then, create a scatter plot to visualize the relationship between home runs (HR) and strikeouts (SO) for these players. Add a line of best fit and appropriate labels to the visualization.

::: callout-tip
Use `inner_join()` to ensure only players with both batting and pitching statistics are included. Use `left_join()` to add player names from the `People` dataset.
:::

```{r}

```

### Question 7

Use `left_join()` to create a dataset that combines information from the `Batting` and `Teams` datasets. Calculate the batting average (total hits divided by total at-bats) for each team and year. Save the result in a new dataset called `team_batting_avg`. Then, create a heatmap to visualize the batting averages by team and year. Filter the dataset to include only years after 2010. Make sure to add appropriate labels and a title to the visualization.

::: callout-tip
Use `geom_tile()` to create the heatmap.
:::

```{r}

```

### Question 8

Use `left_join()` to create a dataset that combines information from the `Batting` and `Fielding` datasets. Calculate the total home runs (HR) for each position (POS) from the years 2000–2023. Save the result in a new dataset called `hr_by_position`. Then, create a bar chart to visualize the total home runs by position, with each position represented by a different colored bar. Make sure to filter for N/A values in player positions and add appropriate labels to the visualization.

::: callout-tip
Use `group_by()` and `summarize()` to calculate total home runs by position.
:::

```{r}

```
